---
title: "Habitat connectivity report"
date: today
params:
  report_data: "report-data.rds"
format:
  html:
    toc: true
    embed-resources: true
  typst:
    toc: true
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(urbioconnect)

report <- read_report_data(params$report_data)

connectivity <- report$connectivity
distances <- report$interpatch_distance

# draw tabs, for each buffer distance, with blank ending line, for pandoc
distance_tabs <- function(draw) {
  cat("\n::: {.panel-tabset}\n")

  purrr::walk(distances, function(distance) {
    cat("\n## ", distance, "m\n\n", sep = "")
    print(draw(distance))
    cat("\n")
  })

  cat("\n:::\n")
}
```

**Species:** `r report$species`

**Interpatch distances:** `r paste0(distances, "m", collapse = ", ")`

**Data resolution:** `r format_resolution(terra::res(report$habitat))` m

Two habitat patches count as connected when the gap between them is no more
than the interpatch distance. Each distance below is a separate analysis of the
same landscape.

# Habitat, interpatch zone and barrier

The habitat, the zone within half an interpatch distance of it, and the
barriers that break connections.

```{r}
#| label: maps-habitat
#| output: asis
distance_tabs(function(distance) {
  gg_barrier_habitat_interpatch_dist(
    barrier = report$barrier,
    buffered = report$buffered_habitat[[as.character(distance)]],
    habitat = report$habitat,
    interpatch_distance = distance,
    species = report$species
  )
})
```

# Connected patches

Habitat coloured by the patch it belongs to. Patches sharing a colour are
connected at this distance.

```{r}
#| label: maps-patches
#| output: asis
distance_tabs(function(distance) {
  plot_patches(
    patch_id = report$patch_id_raster[[as.character(distance)]],
    interpatch_distance = distance,
    species = report$species
  )
})
```

# Connectivity summary

One row per interpatch distance. Species and resolution are the same for every
row, and are given above.

```{r}
#| label: summary-table
# each column is named where it is computed, so a renamed or reordered column
# upstream fails here rather than quietly mislabelling a number
connectivity |>
  dplyr::mutate(
    `Interpatch distance (m)` = interpatch_distance,
    Patches = n_patches,
    `Effective mesh (ha)` = round(effective_mesh_ha, 2),
    `P(connected)` = signif(prob_connectedness, 3),
    `Mean patch area (m2)` = round(patch_area_mean, 1),
    `Total patch area (ha)` = round(patch_area_total_ha, 2),
    .keep = "none"
  ) |>
  knitr::kable()
```

# Largest patches

The ten largest patches at each distance. Every patch is in the summary folder
of the accompanying download.

```{r}
#| label: patch-tables
#| output: asis
patches_by_distance <- rlang::set_names(
  patch_sizes(connectivity),
  as.character(distances)
)

distance_tabs(function(distance) {
  patches_by_distance[[as.character(distance)]] |>
    dplyr::arrange(dplyr::desc(area)) |>
    head(10) |>
    dplyr::mutate(
      `Patch ID` = patch_id,
      `Area (m2)` = round(area, 1),
      .keep = "none"
    ) |>
    knitr::kable(row.names = FALSE)
})
```

```{r}
#| label: over-distance
#| output: asis
#| fig-height: !expr urbioconnect::urbio_figure_size()$tall_height
# a single distance is a single point, so there is nothing to plot
if (length(distances) > 1) {
  cat("# Change over distance\n\n")
  cat("How the metrics move as the interpatch distance grows.\n\n")
  print(plot_connectivity(connectivity))
}
```

# The spatial files

The download accompanying this report holds the patch raster as a GeoTIFF and
the patch polygons as both a GeoPackage and a shapefile, ready to open in QGIS
or ArcGIS, for each interpatch distance. Its `README.md` lists every file.
Patch areas are in square metres.
